
% ---------------------------------------------------------------
% A.1 — Series glossary
% ---------------------------------------------------------------

\subsection{Series Glossary}

\begin{table}[!ht]
	\centering
	\small
	\setlength{\tabcolsep}{3pt}
	\renewcommand{\arraystretch}{1.15}
	\caption{Replication series: symbols, sources, and transformations}
	\label{tab:app_glossary}
	\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.20\textwidth}
			>{\raggedright\arraybackslash}p{0.26\textwidth}
			>{\raggedright\arraybackslash}p{0.28\textwidth}
			>{\raggedright\arraybackslash}p{0.18\textwidth}@{}}
		\hline
		Symbol / Code & Variable & Source (table, line) & Transformation \\
		\hline
		$\text{GVAcorp}_t$ 
		& Corporate gross value added (corrected)
		& NIPA 1.14 (L1) + imputed-interest adj.\ (Table~7.11)
		& nominal \\[3pt]
		$\text{KGCcorp}_t$ 
		& Corporate gross capital stock (GPIM-adjusted)
		& Fixed Asset 6.1 (L9); IRS Statistics of Income; GPIM recursion
		& nominal \\[3pt]
		$\text{KNCcorp}_t$ 
		& Corporate net capital stock (GPIM-adjusted)
		& same as above
		& nominal \\[3pt]
		$\text{KNCcorpbea}_t$
		& Corporate net capital stock (BEA 2011 official)
		& Fixed Asset 6.1 (L9)
		& nominal \\[3pt]
		$p_t$ 
		& Common deflator
		& $pKN$ index, 2005=100 
		& (KNCcorp/KNRcorp)*100 \\[3pt]
		$y_t$
		& Real log output
		& constructed
		& $\ln(\text{VAcorp}_t/p_t)$ \\[3pt]
		$k_t$
		& Real log capital
		& constructed
		& $\ln(\text{KGCcorp}_t/p_t)$ \\[3pt]
		$\text{Profshcorp}_t$ 
		& Corporate profit share (corrected)
		& NIPA 1.14 (L8, L1$-$L2) + imputed-interest adj.
		& ratio \\
		\hline
	\end{tabular}
	\par\vspace{0.4em}
	\footnotesize\textit{Note:} All BEA and NIPA data downloaded in the
	replication vintage. Shaikh's original series used data downloaded in
	2011 \citep[Appendix~6.7, fn.~1]{Shaikh2016}; discrepancies attributable
	to subsequent BEA revisions are possible and noted where material.
	\par\vspace{0.2em}
	\footnotesize\textit{Source:}
	\href{https://www.realecon.org/data}{Real Economic Analysis Data Tables}
\end{table}

% ---------------------------------------------------------------
% A.2 — Imputed-interest correction
% ---------------------------------------------------------------

\subsection{Imputed-Interest Correction}

The NIPA corporate value added figure includes imputed financial
intermediation services as a cost to non-financial corporations. The
corporate imputed-interest adjustment removes this component:
%
\begin{equation}
	\label{eq:app_impint}
	\text{CorpImpIntAdj}_t
	\;\equiv\;
	-\,\text{BankMonIntPaid}_t
	\;-\; \text{CorpNFNetImpIntPaid}_t,
\end{equation}
%
where $\text{BankMonIntPaid}_t$ is NIPA Table~7.11
(lines 4+44+73 minus lines 28+52+91) and
$\, \, \text{CorpNFNetImpIntPaid}_t$ is NIPA Table~7.11 (line~74 minus line~53).
In 2009 the adjustment raises corporate net operating surplus by
approximately 10\,\%.

% ---------------------------------------------------------------
% A.3 — Corporate output: GVAcorp
% ---------------------------------------------------------------

\subsection{Corrected Corporate Output: \texorpdfstring{$\text{GVAcorp}_t$}{GVAcorp}}

\begin{equation}
	\label{eq:app_gva}
	\text{GVAcorp}_t
	\;\equiv\;
	\text{GVAcorpnipa}_t \;+\; \text{CorpImpIntAdj}_t,
\end{equation}
%
where $\text{GVAcorpnipa}_t$ is NIPA Table~1.14 (line~1) and
$\text{CorpImpIntAdj}_t$ is defined in equation~\eqref{eq:app_impint}.

% ---------------------------------------------------------------
% A.4 — Corporate capital stock: KGCcorp (GPIM)
% ---------------------------------------------------------------

\subsection{Reconstructed Corporate Capital Stock: \texorpdfstring{$\text{KGCcorp}_t$}{KGCcorp}}

\paragraph{GPIM accumulation rule.}
The current-cost capital stock is accumulated as:
%
\begin{equation}
	\label{eq:app_gpim}
	K_t \;=\; IG_t \;+\; z_t^* \cdot K_{t-1},
	\qquad
	z_t^* \;\equiv\; (1 - z_t)\,\frac{p^K_t}{p^K_{t-1}},
\end{equation}
%
where $IG_t$ is nominal gross investment, $z_t$ is the period depreciation
rate, and $z_t^*$ is the survival-and-revaluation factor that reflates the
surviving stock to current replacement cost.

\paragraph{Three adjustments (applied before the recursion).}

\begin{enumerate}
	\item \textbf{Depletion rates.} BEA 1993 finite service lives replace
	the BEA 2011 infinite-lives assumption, implying higher depreciation
	and lower retirement rates consistent with physical exhaustion.
	
	\item \textbf{Initial values.} BEA 1993 initial values anchored to
	1925 replace the BEA 2011 starting values. The 1993 values are lower,
	producing faster accumulation growth and convergence to the official
	path by approximately 1977.
	
	\item \textbf{Great Depression correction.} The IRS corporate book
	value index \citep[Series~V~115, pp.~924--926]{UScensus1975} is applied
	over 1925--1947, rescaling the BEA historical stock by the IRS/BEA
	ratio each year. The GPIM recursion resumes from 1947. Net effect:
	$\text{KNCcorp}_t$ is approximately 28\,\% lower in 1947 than the BEA
	2011 official measure.
\end{enumerate}

\paragraph{Inventories.}
IRS corporate balance sheet inventory ratios are extrapolated over
1947--1989 via a linear fit to the 1990--2011 IRS/BEA ratio, rescaled to
the adjusted GPIM stocks, and added to $\text{KGCcorp}_t$. Source:
IRS Statistics of Income 1926--2011; FRB Flow of Funds FL105015205.A for
non-financial corporate inventories in current cost.

\paragraph{Accuracy.}
The average ratio of the GPIM approximation to the BEA current-cost stock
over 1947--2005 is 99.60\,\%.

\noindent Full recursion inputs and depletion-rate construction are in
\citet[Appendix~6.7, Section~V; Appendix Table~6.8.II.3]{Shaikh2016}.

% ---------------------------------------------------------------
% A.5 — Common deflator
% ---------------------------------------------------------------

\subsection{Common Deflator: \texorpdfstring{$p_t$}{p\_t}}

The replication uses the $pKN$ implicit price deflator based at $100 = 2005$ from current prices net capital stocks as $p_t$. Esimtated by Shaikh under GPIM rules in detriment of quality adjusted chain weights published in current (and 2011 also) BEA releases of Fixed Assets tables in order to sustain stock-flow consistency. 

% ---------------------------------------------------------------
% A.6 — Corrected corporate profit share: Profshcorp
% ---------------------------------------------------------------

\subsection{Corrected Corporate Profit Share: \texorpdfstring{$\text{Profshcorp}_t$}{Profshcorp}}

The same imputed-interest adjustment applied to $\text{GVAcorp}_t$ is
applied to both the numerator and denominator of the profit share:
%
\begin{align}
	\text{NOScorp}_t
	&\;\equiv\; \text{NOScorpnipa}_t \;+\; \text{CorpImpIntAdj}_t,
	\label{eq:app_nos} \\
	\text{VAcorp}_t
	&\;\equiv\; \text{VAcorpnipa}_t \;+\; \text{CorpImpIntAdj}_t,
	\label{eq:app_va} \\
	\text{Profshcorp}_t
	&\;\equiv\; \frac{\text{NOScorp}_t}{\text{VAcorp}_t},
	\label{eq:app_profshcorp}
\end{align}
%
where $\text{NOScorpnipa}_t$ is NIPA Table~1.14 (line~8) and
$\text{VAcorpnipa}_t$ is NIPA Table~1.14 (line~1 minus line~2).